Convergence rate of MAP estimates \\ for the exponential family

### 2020-11-27 main.tex
Move mirror descent forward and stuff to appendix

@@

\subsubsection{Three point lemma}
These proofs rely on another basic property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
\begin{align}
	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
\end{align}
where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector. Then 
\begin{align}
	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
\end{align}
This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).

@@

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$. 
\begin{align}
	\Cov(g_t,\natp_{t+1}) 
	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \lr_t \sigma^2, \forall t \\
	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
\begin{align}
	g_t - \nabla f(\natp_t) = \E[T(X)] - T(X_{t+1})	
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanconj} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.

@@

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    \label{eq:bregmanconj}
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ? 
\begin{align}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right )
\end{align}

@@

\subsection{Fenchel conjugate motivation to self-concordance}

@@

\newpage

>>>>

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
\end{align}

@@

@@

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$. 
\begin{align}
	\Cov(g_t,\natp_{t+1}) 
	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \lr_t \sigma^2, \forall t \\
	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
\begin{align}
	g_t - \nabla f(\natp_t) = \E[T(X)] - T(X_{t+1})	
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.

@@

@@

@@

\newpage
\appendix

\section{Three point lemma}
Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
\begin{align}
	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
\end{align}
where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector. Then 
\begin{align}
	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
\end{align}
This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).

\section{Fenchel conjugate motivation to self-concordance}
